DOpElib:\\ Differential Equations and Optimization Environment

### 2013-12-12 Examples/Documentation/foreword.tex
Update Doku in Einklang mit Christians und Winnis Bemerkungen/Zustimmung

\chapter{Foreword}
\textcolor{red}{Ich habe Section 2.8 etc. 'how this library is structured' zu 
einem eigenen chapter gemacht}
In this report, we describe the \dope{}
\textit{Differential Equations and  Optimization Environment} project. 
Originally, the project was initiated in the year 2009 at the 
University of Heidelberg (Germany) in the numerical analysis
group of Rolf Rannacher.

\textcolor{blue}{Eigentlich kann dieser ganze paragraph weg. Im naechsten
  Paragraphen steht das
im Prinzip nochmal!
Its key feature is to give a unified interface to high level algorithms such as 
time stepping methods, nonlinear solvers and optimization routines. 
We aim that the user should only need to write those parts
of code that are problem dependent while all invariant parts of the algorithms
should be reusable without any need for further coding.
In particular, the user should be able to switch between various different 
algorithms without the need to rewrite the problem dependent code, though he or she will
have to replace the algorithm object with an other one. 
Another aim of \dope{} is to provide a software toolkit to solve forward PDE
problems as well as optimal control problems constrained by PDE. The
solution of a broad variety of PDE is possible in deal.II as well, but
\dope{} concentrates on a unified approach for both linear and nonlinear
problems by interpreting every PDE problem as nonlinear and applying a
Newton method to solve it. 
The focus is on the numerical solution of both stationary and nonstationary
problems which come from different application fields, like elasticity and
plasticity, fluid dynamics, and multiphysics problems such as 
fluid-structure interactions.
}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
At the present stage the following features are supported by the library
\begin{itemize}
\item Solution of stationary and nonstationary PDEs in 1d, 2d, and 3d.
\item Various time stepping schemes (based on finite differences), 
  such as forward Euler, backward Euler,
  Crank-Nicolson, shifted Crank-Nicolson, and Fractional-Step-$\Theta$ scheme.
\item All finite elements of from deal.II including hp-support.
\item Several examples showing the solution of several PDEs including
   Poisson, Navier-Stokes, plasticity, fluid-structure interaction problems,
and finally a coupled Biot-Lam\'e-Navier system.
\item Self-written line search and trust region newton algorithms for the 
   solution of optimization problems with PDEs \cite{NoWr00}
\item Interface to SNOPT for the solution of optimization problems with PDEs and
  additional other constraints.
\item Several examples showing how to solve various kinds of optimization problems
  involving stationary PDE constraints.
\item Mesh adaptation and \textcolor{red}{goal-oriented error estimation}.
\item Different spatial triangulations for control and state variables.
\end{itemize}

@@

\paragraph{Contributors \& Thanks:}
\textcolor{red}{Ich habe das von oben hierhin verschoben und etwas 
umgeschrieben:
The \dope{} project is 
mainly based on the \textit{deal.II} finite element library which has been developed
 initially by W. Bangerth, R. Hartmann, and G. Kanschat \cite{deal} and 
now maintained by \cite{deal2}. Special thanks go to them!
In addition to deal.II, the authors 
gratefully acknowledge their past experience as well as discussions with 
the authors of  
the software toolkits \texttt{Gascoigne} and \texttt{RoDoBo}
\cite{Gascoigne} and \cite{rodobo-web}, 
from which some of the ideas to modularize the algorithms have arisen.
Similar thanks go to the developers of \texttt{Ipopt},~\cite{ipopt}.
Last, but not least, we would like to express our gratitude 
to all contributors listed in Section~\ref{sec:contrib}
for their respective contribution to the library.
}

>>>>

\chapter{Foreword}
In this report, we describe the \dope{}
\textit{Differential Equations and  Optimization Environment} project. 
Originally, the project was initiated in the year 2009 at the 
University of Heidelberg (Germany) in the numerical analysis
group of Rolf Rannacher.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
At the present stage the following features are supported by the library
\begin{itemize}
\item Solution of stationary and nonstationary PDEs in 1d, 2d, and 3d.
\item Various time stepping schemes (based on finite differences), 
  such as forward Euler, backward Euler,
  Crank-Nicolson, shifted Crank-Nicolson, and Fractional-Step-$\Theta$ scheme.
\item All finite elements of from deal.II including hp-support.
\item Several examples showing the solution of several PDEs including
   Poisson, Navier-Stokes, plasticity, fluid-structure interaction problems,
and finally a coupled Biot-Lam\'e-Navier system.
\item Self-written line search and trust region newton algorithms for the 
   solution of optimization problems with PDEs \cite{NoWr00}
\item Interface to SNOPT for the solution of optimization problems with PDEs and
  additional other constraints.
\item Several examples showing how to solve various kinds of optimization problems
  involving stationary PDE constraints.
\item Mesh adaptation and goal-oriented error estimation with
the dual-weighted residual (DWR) method.
\item Different spatial triangulations for control and state variables.
\end{itemize}

@@

%\paragraph{Contributors \& Thanks:}
\paragraph{Thanks:}
The \dope{} project is 
mainly based on the \textit{deal.II} finite element library which has been developed
 initially by W. Bangerth, R. Hartmann, and G. Kanschat \cite{deal} and 
now maintained by \cite{deal2}. Special thanks go to them!

In addition to deal.II, the authors 
gratefully acknowledge their past experience as well as discussions with 
the authors of  
the software toolkits \texttt{Gascoigne} and \texttt{RoDoBo}
\cite{Gascoigne} and \cite{rodobo-web}, 
from which some of the ideas to modularize the algorithms have arisen.
Similar thanks go to the developers of \texttt{Ipopt},~\cite{ipopt}.

Last, but not least, we would like to express our gratitude 
to all contributors listed in Section~\ref{sec:contrib}
for their respective contribution to the library.

### 2013-12-12 Examples/Documentation/tests.tex
typo
Update Doku in Einklang mit Christians und Winnis Bemerkungen/Zustimmung

\chapter{Regression tests}\label{chap:test}
\textcolor{red}{Ich habe die Ueberschrift geandert}
The DOpE test suite consists of regression tests. They are run to compare 
the output to previous outputs. This is useful (necessary) after 
changing programming code anywhere in the library. If a test
succeeds, everything is fine in the library. If not, you should not
check in your code into DOpE. Please make sure what is going wrong and WHY!

Every command is computed via a Makefile. In the basic example 
directory.

\textcolor{red}{
\begin{remark}
Please keep in mind that there are two optimization examples that 
need external packages to be able to run successfully. These tests will
fail if these libraries are not installed!
\end{remark}
}

>>>>

\chapter{Regression Tests}\label{chap:test}
The \dope{} test suite consists of regression tests. They are run to compare 
the output to previous outputs. This is useful (necessary) after 
changing programming code anywhere in the library.

\begin{itemize}
\item If a test
succeeds, everything is fine in the library. 
\item If not, you should not
check in your code into \dope{}. 
Please make sure what is going wrong and WHY!
\item 
Every command is computed via a Makefile in the basic example 
directory.
\end{itemize}

\begin{remark}
Please keep in mind that there are two optimization examples that 
need external packages to be able to run successfully. These tests will
fail if these libraries are not installed!
\end{remark}

### 2013-12-12 Examples/OPT/StatPDE/Example1/content.tex
Doku anpassung
Update Doku in Einklang mit Christians und Winnis Bemerkungen/Zustimmung

In abstract form, we are given the following optimization problem:
\begin{equation*}
J(q,u) \, \rightarrow \, \min , \quad a(q,u)(\psi) = 0 \quad \forall\psi\in V
\end{equation*}
Lagrangian:
\begin{equation*}
{\cal L}(q,u,z):=J(q,u) - a(q,u)(z)
\end{equation*}
Optimality system (KKT system):
\begin{align*}
a_u'(q,u)(\phi , z) &= J'_u(q,u)(\phi) \quad \forall \phi\in V \\
a_q'(q,u)(\chi , z) &= J'_q(q,u)(\chi) \quad \forall \chi\in Q \\
a(q,u)(\psi) &= 0 \qquad \forall\psi\in V
\end{align*}
or equivalently, in terms of the Lagrangian
\begin{align*}
{\cal L}'_u(q,u,z)(\phi) &= 0 \quad\forall \phi\in V \quad\text{(Adjoint Equation)} \\
{\cal L}'_q(q,u,z)(\chi) &= 0 \quad\forall \chi\in V \quad\text{(Gradient Equation)} \\
{\cal L}'_z(q,u,z)(\psi) &= 0 \quad\forall \psi\in V \quad\text{(State Equation)}
\end{align*}
The continuous problem is discretized by a standard Galerkin method using 
finite dimensional subspaces $Q_h \times V_h \subset Q\times V$:
\begin{equation*}
J(q,u) \, \rightarrow \, \min , \quad a(q,u)(\psi) = 0 \quad \forall\psi\in V
\end{equation*}
Discrete saddle-point problems
\begin{align*}
a_u'(q_h,u_h)(\phi_h , z_h) &= J'_u(q_h,u_h)(\phi_h) \quad \forall \phi_h\in V_h \\
a_q'(q_h,u_h)(\chi_h , z_h) &= J'_q(q_h,u_h)(\chi_h) \quad \forall \chi_h\in Q_h \\
a(q_h,u_h)(\psi_h) &= 0 \qquad \forall\psi_h\in V_h
\end{align*}
\subsubsection{Solution process}
In this section, we briefly discuss the solution process for the 
optimization problem. For further details, we refer to the standard literature.
The unconstrained optimal control problem is reformulated as follows. 
We introduce the solution operator $S:Q\rightarrow V$ of the state equation.
Then:
\begin{equation*}
j(q):= J(q,S(q))\, \rightarrow \, \min , \quad a(q,S(q))(\Psi) = 0 \quad 
\forall \Psi \in V.
\end{equation*}
The local existence and sufficient regularity of $S$ is assumed. The 
necessary optimality conditions of first and second order are
\begin{equation*}
j'(q)(\delta q) = 0 , \quad j''(q)(\delta q,\delta q) \geq 0 \quad 
\forall\delta q\in Q.
\end{equation*} 
The derivatives of the reduced functional can be computed using the 
Lagrangian
\begin{equation*}
{\cal L}(q,u,z) =J(q,u) - a(q,u)(z)
\end{equation*}
as already introduced. Let $q\in Q$, and the corresponding state $u=S(q)\in V$ 
be given. To calculate the derivative of the reduced cost functional $j$, 
we introduce the dual variable
$z\in V$ solving the\\[3mm]
%
\textbf{Dual equation}\\
\begin{equation*}
{\cal L}_U'(q,U,Z)(\Psi) = 0 \quad\forall\Psi\in V.
\end{equation*}
Then,
\begin{equation*}
j'(q)(\delta q) = {\cal L}_q' (q,U,Z)(\delta q) \quad \text{for }\delta q\in Q.
\end{equation*}
\\[2mm]
%
To calculate the second derivatives, we need to solve additional equations.\\
%
Let $\delta q\in Q$ be a given direction. Then we search $\delta U\in V$  
solving the\\[2mm]
%
\textbf{Tangent equation}\\
%
\begin{equation*}
{\cal L}_{qZ}'' (q,U,Z)(\delta q,\Phi) + {\cal L}_{UZ}'' (q,U,Z)(\delta
U,\Phi) = 0
\quad\forall \Phi\in V.
\end{equation*}
Further, we have an auxilliary\\[2mm]  
%
\textbf{Dual for Hessian equation}
to find $\delta Z\in V$ solving
\begin{equation*}
{\cal L}_{qU}'' (q,U,Z)(\delta q,\Phi) + {\cal L}_{UU}'' 
(q,U,Z)(\delta U,\Phi) + {\cal L}_{ZU}'' (q,U,Z)(\delta Z,\Phi) =0 
\quad\forall \Phi\in V.
\end{equation*}

Then, for $\delta r\in Q$, we can express the second derivatives of $j$ by
\begin{align*}
j''(q)(\delta q, \delta r) 
&= {\cal L}_{qq}'' (q,U,Z)(\delta q,\delta r) \\
&\;\;\;\;+ {\cal L}_{Uq}'' (q,U,Z)(\delta U,\delta r) \\ 
&\;\;\;\;+ {\cal L}_{Zq}'' (q,U,Z)(\delta Z,\delta r).
\end{align*}
With these terms, we can calculate the Newton direction $\delta q$, at a given 
iterate $q^n$, as solution to the problem
\begin{align*}
j''(q^n )(\delta q,\chi ) =&\, -j'(q^n)(\chi) \quad\forall\chi \in Q.
\end{align*}

@@

To solve the linear equations the following matrices are needed{\color{red}
\begin{align*}
&\texttt{ElementMatrix} 
&\Leftrightarrow \quad a_{i,j} &= a'_u(q,u)(\phi_j,\phi_i), \\
&\texttt{ControlElementMatrix}  
&\Leftrightarrow \quad a_{i,j} &= (\phi^q_j, \phi^q_i)_Q.
\end{align*}}
The first one is required for all primal and dual PDE solves, while the second one is needed 
to calculate the Riesz representation of the derivatives of $j$. 
If desired, the matrix for the adjoint PDEs can be provided separately as 
\texttt{ElementMatrix\underline{ }T}, but otherwise this will be 
calculated automatically from the primal matrix.

Additional terms are needed to calculate the corresponding 
right hand sides. These are for the PDE the following: 
{\color{red}
\begin{align*}
&\texttt{ElementEquation}\text{ (state)} &\Leftrightarrow &\quad a(q,u)(\phi_i), \\
&\texttt{ElementRightHandSide}\text{ (state)}  &\Leftrightarrow &\quad f(\phi_i), \\
&\texttt{ControlElementEquation}\text{ (gradient or hessian)}  &\Leftrightarrow &\quad (\nabla j(q),\phi^q_i)_Q,
\end{align*}
as well as 
\begin{align*}
&\texttt{ElementEquation\underline{ }U}\text{ (adjoint)} 
&\Leftrightarrow &\quad a_u'(q,u)(\phi_i, z),\\
&\texttt{ElementEquation\underline{ }Q}\text{ (gradient)} 
&\Leftrightarrow &\quad a_q'(q,u)(\phi^q_i,z), \\
\end{align*}
the terms
\begin{align*}
&\texttt{ElementEquation\underline{ }UU}\text{ (adjoint hessian)}  
&\Leftrightarrow &\quad a_{uu}''(q,u)(\delta u, \phi_i, z),\\
&\texttt{ElementEquation\underline{ }UQ}\text{ (hessian)}
&\Leftrightarrow &\quad a_{uq}''(q,u)(\delta u, \phi^q_i, z),\\
&\texttt{ElementEquation\underline{ }QU}\text{ (adjoint hessian)} 
&\Leftrightarrow &\quad a_{qu}''(q,u)(\delta q, \phi_i, z), \\
&\texttt{ElementEquation\underline{ }QQ}\text{ (hessian)} 
&\Leftrightarrow &\quad a_{qq}''(q,u)(\delta q, \phi^q_i, z),\\
&\texttt{ElementEquation\underline{ }UT}\text{ (tangent)}  
&\Leftrightarrow &\quad a_u'(q,u)(\delta u, \phi_i), \\
&\texttt{ElementEquation\underline{ }QT}\text{ (tangent)}  
&\Leftrightarrow &\quad a_q'(q,u)(\delta q, \phi_i),
\end{align*}
and finally
\begin{align*}
&\texttt{ElementEquation\underline{ }UTT}\text{ (adjoint hessian)} \quad 
&\Leftrightarrow &\quad  a_u'(q,u)(\phi_i, \delta z), \\
&\texttt{ElementEquation\underline{ }QTT}\text{ (hessian)} \quad 
&\Leftrightarrow &\quad a_{q}'(q,u)(\phi^q_i,\delta z).
\end{align*}}
As for PDE problems, it is up to the user to decide if the \texttt{ElementRightHandSide}
is used, or if the terms are included in the \texttt{ElementEquation}.

For the cost functional, we have to provide{\color{red}
\begin{align*}
&\texttt{ElementValue}\text{ (all)}   &\Leftrightarrow &\quad J(q,u), \\
&\texttt{ElementValue\underline{ }U}\text{ (all)}    &\Leftrightarrow &\quad J_u'(q,u)(\phi_i), \\
&\texttt{ElementValue\underline{ }Q}\text{ (all)}    &\Leftrightarrow &\quad J_q'(q,u)(\phi_i), \\
&\texttt{ElementValue\underline{ }UU}\text{ (all)}  &\Leftrightarrow &\quad J_{uu}''(q,u)(\delta u, \phi_i), \\
&\texttt{ElementValue\underline{ }UQ}\text{ (all)}  &\Leftrightarrow &\quad J_{uq}''(q,u)(\delta u, \phi^q_i), \\
&\texttt{ElementValue\underline{ }QU}\text{ (all)}   &\Leftrightarrow &\quad J_{qu}''(q,u)(\delta q, \phi_i), \\
&\texttt{ElementValue\underline{ }QQ}\text{ (all)}  &\Leftrightarrow &\quad J_{qq}''(q,u)(\delta q, \phi^q_i).
\end{align*}}

>>>>

In abstract form, we are given the following optimization problem:
\begin{equation*}
J(q,u) \, \rightarrow \, \min , \quad a(q,u)(\psi) = 0 \quad \forall\psi\in V
\end{equation*}
Lagrangian:
\begin{equation*}
{\cal L}(q,u,z):=J(q,u) - a(q,u)(z)
\end{equation*}
Optimality system (KKT system):
\begin{align*}
a_u'(q,u)(\phi , z) &= J'_u(q,u)(\phi) \quad \forall \phi\in V \\
a_q'(q,u)(\chi , z) &= J'_q(q,u)(\chi) \quad \forall \chi\in Q \\
a(q,u)(\psi) &= 0 \qquad \forall\psi\in V
\end{align*}
or equivalently, in terms of the Lagrangian
\begin{align*}
{\cal L}'_u(q,u,z)(\phi) &= 0 \quad\forall \phi\in V \quad\text{(Adjoint Equation)} \\
{\cal L}'_q(q,u,z)(\chi) &= 0 \quad\forall \chi\in V \quad\text{(Gradient Equation)} \\
{\cal L}'_z(q,u,z)(\psi) &= 0 \quad\forall \psi\in V \quad\text{(State Equation)}
\end{align*}
The continuous problem is discretized by a standard Galerkin method using 
finite dimensional subspaces $Q_h \times V_h \subset Q\times V$:
\begin{equation*}
J(q,u) \, \rightarrow \, \min , \quad a(q,u)(\psi) = 0 \quad \forall\psi\in V
\end{equation*}
Discrete saddle-point problems
\begin{align*}
a_u'(q_h,u_h)(\phi_h , z_h) &= J'_u(q_h,u_h)(\phi_h) \quad \forall \phi_h\in V_h \\
a_q'(q_h,u_h)(\chi_h , z_h) &= J'_q(q_h,u_h)(\chi_h) \quad \forall \chi_h\in Q_h \\
a(q_h,u_h)(\psi_h) &= 0 \qquad \forall\psi_h\in V_h
\end{align*}
\subsubsection{Solution process}
In this section, we briefly discuss the solution process for the 
optimization problem. For further details, we refer to the standard literature.
The unconstrained optimal control problem is reformulated as follows. 
We introduce the solution operator $S:Q\rightarrow V$ of the state equation.
Then:
\begin{equation*}
j(q):= J(q,S(q))\, \rightarrow \, \min , \quad a(q,S(q))(\Psi) = 0 \quad 
\forall \Psi \in V.
\end{equation*}
The local existence and sufficient regularity of $S$ is assumed. The 
necessary optimality conditions of first and second order are
\begin{equation*}
j'(q)(\delta q) = 0 , \quad j''(q)(\delta q,\delta q) \geq 0 \quad 
\forall\delta q\in Q.
\end{equation*} 
The derivatives of the reduced functional can be computed using the 
Lagrangian
\begin{equation*}
{\cal L}(q,u,z) =J(q,u) - a(q,u)(z)
\end{equation*}
as already introduced. Let $q\in Q$, and the corresponding state $u=S(q)\in V$ 
be given. To calculate the derivative of the reduced cost functional $j$, 
we introduce the dual variable
$z\in V$ solving the\\[3mm]
%
\textbf{Dual equation}\\
\begin{equation*}
{\cal L}_u'(q,u,z)(\psi) = 0 \quad\forall\psi\in V.
\end{equation*}
Then,
\begin{equation*}
j'(q)(\delta q) = {\cal L}_q' (q,u,z)(\delta q) \quad \text{for }\delta q\in Q.
\end{equation*}
\\[2mm]
%
To calculate the second derivatives, we need to solve additional equations.\\
%
Let $\delta q\in Q$ be a given direction. Then we search $\delta u\in V$  
solving the\\[2mm]
%
\textbf{Tangent equation}\\
%
\begin{equation*}
{\cal L}_{qz}'' (q,u,z)(\delta q,\phi) + {\cal L}_{uz}'' (q,u,z)(\delta
u,\phi) = 0
\quad\forall \phi\in V.
\end{equation*}
Further, we have an auxilliary\\[2mm]  
%
\textbf{Dual for Hessian equation}
to find $\delta z\in V$ solving
\begin{equation*}
{\cal L}_{qu}'' (q,u,z)(\delta q,\phi) + {\cal L}_{uu}'' 
(q,u,z)(\delta u,\phi) + {\cal L}_{zu}'' (q,u,z)(\delta z,\phi) =0 
\quad\forall \phi\in V.
\end{equation*}

Then, for $\delta r\in Q$, we can express the second derivatives of $j$ by
\begin{align*}
j''(q)(\delta q, \delta r) 
&= {\cal L}_{qq}'' (q,u,z)(\delta q,\delta r) \\
&\;\;\;\;+ {\cal L}_{uq}'' (q,u,z)(\delta u,\delta r) \\ 
&\;\;\;\;+ {\cal L}_{zq}'' (q,u,z)(\delta z,\delta r).
\end{align*}
With these terms, we can calculate the Newton direction $\delta q$, at a given 
iterate $q^n$, as solution to the problem
\begin{align*}
j''(q^n )(\delta q,\chi ) =&\, -j'(q^n)(\chi) \quad\forall\chi \in Q.
\end{align*}

@@

To solve the linear equations the following matrices are needed
\begin{align*}
&\texttt{ElementMatrix} 
&\Leftrightarrow \quad a_{i,j} &= a'_u(q,u)(\phi_j,\phi_i), \\
&\texttt{ControlElementMatrix}  
&\Leftrightarrow \quad a_{i,j} &= (\phi^q_j, \phi^q_i)_Q.
\end{align*}
The first one is required for all primal and dual PDE solves, while the second one is needed 
to calculate the Riesz representation of the derivatives of $j$. 
If desired, the matrix for the adjoint PDEs can be provided separately as 
\texttt{ElementMatrix\underline{ }T}, but otherwise this will be 
calculated automatically from the primal matrix.

Additional terms are needed to calculate the corresponding 
right hand sides. These are for the PDE the following: 
{
\begin{align*}
&\texttt{ElementEquation}\text{ (state)} &\Leftrightarrow &\quad a(q,u)(\phi_i), \\
&\texttt{ElementRightHandSide}\text{ (state)}  &\Leftrightarrow &\quad f(\phi_i), \\
&\texttt{ControlElementEquation}\text{ (gradient or hessian)}  &\Leftrightarrow &\quad (\nabla j(q),\phi^q_i)_Q,
\end{align*}
as well as 
\begin{align*}
&\texttt{ElementEquation\underline{ }U}\text{ (adjoint)} 
&\Leftrightarrow &\quad a_u'(q,u)(\phi_i, z),\\
&\texttt{ElementEquation\underline{ }Q}\text{ (gradient)} 
&\Leftrightarrow &\quad a_q'(q,u)(\phi^q_i,z), \\
\end{align*}
the terms
\begin{align*}
&\texttt{ElementEquation\underline{ }UU}\text{ (adjoint hessian)}  
&\Leftrightarrow &\quad a_{uu}''(q,u)(\delta u, \phi_i, z),\\
&\texttt{ElementEquation\underline{ }UQ}\text{ (hessian)}
&\Leftrightarrow &\quad a_{uq}''(q,u)(\delta u, \phi^q_i, z),\\
&\texttt{ElementEquation\underline{ }QU}\text{ (adjoint hessian)} 
&\Leftrightarrow &\quad a_{qu}''(q,u)(\delta q, \phi_i, z), \\
&\texttt{ElementEquation\underline{ }QQ}\text{ (hessian)} 
&\Leftrightarrow &\quad a_{qq}''(q,u)(\delta q, \phi^q_i, z),\\
&\texttt{ElementEquation\underline{ }UT}\text{ (tangent)}  
&\Leftrightarrow &\quad a_u'(q,u)(\delta u, \phi_i), \\
&\texttt{ElementEquation\underline{ }QT}\text{ (tangent)}  
&\Leftrightarrow &\quad a_q'(q,u)(\delta q, \phi_i),
\end{align*}
and finally
\begin{align*}
&\texttt{ElementEquation\underline{ }UTT}\text{ (adjoint hessian)} \quad 
&\Leftrightarrow &\quad  a_u'(q,u)(\phi_i, \delta z), \\
&\texttt{ElementEquation\underline{ }QTT}\text{ (hessian)} \quad 
&\Leftrightarrow &\quad a_{q}'(q,u)(\phi^q_i,\delta z).
\end{align*}}
As for PDE problems, it is up to the user to decide if the \texttt{ElementRightHandSide}
is used, or if the terms are included in the \texttt{ElementEquation}.

For the cost functional, we have to provide{
\begin{align*}
&\texttt{ElementValue}\text{ (all)}   &\Leftrightarrow &\quad J(q,u), \\
&\texttt{ElementValue\underline{ }U}\text{ (all)}    &\Leftrightarrow &\quad J_u'(q,u)(\phi_i), \\
&\texttt{ElementValue\underline{ }Q}\text{ (all)}    &\Leftrightarrow &\quad J_q'(q,u)(\phi_i), \\
&\texttt{ElementValue\underline{ }UU}\text{ (all)}  &\Leftrightarrow &\quad J_{uu}''(q,u)(\delta u, \phi_i), \\
&\texttt{ElementValue\underline{ }UQ}\text{ (all)}  &\Leftrightarrow &\quad J_{uq}''(q,u)(\delta u, \phi^q_i), \\
&\texttt{ElementValue\underline{ }QU}\text{ (all)}   &\Leftrightarrow &\quad J_{qu}''(q,u)(\delta q, \phi_i), \\
&\texttt{ElementValue\underline{ }QQ}\text{ (all)}  &\Leftrightarrow &\quad J_{qq}''(q,u)(\delta q, \phi^q_i).
\end{align*}}

### 2013-12-12 Examples/OPT/StatPDE/Example3/content.tex
Update Doku. Habe Winnis-Superfunktionen fuer den Moment auskommentiert, da ich trotz tex-Update sonst nicht kompilieren kann
Update Doku in Einklang mit Christians und Winnis Bemerkungen/Zustimmung

\textbf{Remark:} Since, we use the 
symmetric stress tensor, we need to subtract the nonsymmetric part 
on the outflow boundary, related to the do-nothing condition.
$\diamond$

@@

\twick{
\subsubsection{Program description}
The implementation of this example does not introduce any new \dope{}-specific
features but shows that more complicated equations such as the Navier-Stokes 
system can be used as forward problem for the optimization process. 
This example builds on the previous Example \ref{OPT_Stat_Param_Lin_Ellipt}.
}

>>>>

\begin{remark} Since, we use the 
symmetric stress tensor, we need to subtract the nonsymmetric part 
on the outflow boundary, related to the do-nothing condition.
$\diamond$
\end{remark}

@@

\subsubsection{Program description}
The implementation of this example does not introduce any new \dope{}-specific
features but shows that more complicated equations such as the Navier-Stokes 
system can be used as forward problem for the optimization process. 
This example builds on the previous Example \ref{OPT_Stat_Param_Lin_Ellipt}.

### 2013-12-12 Examples/OPT/StatPDE/Example6/content.tex
Update Doku in Einklang mit Christians und Winnis Bemerkungen/Zustimmung

In this example, we introduce the handling of \textit{local} constraints
whereas the mixture of local and \textit{global} constraints will be discussed
in the Example \ref{OPT_Stat_MBB-Beam}. 
%
First, we implement the upper and lower control bounds in 
\texttt{localconstraints.h}, i.e,
\[
q_{\min} \leq q \leq q_{\max}.
\]
with $q_{\min} = -500$ and $q_{\max} = 500$.
The constraints are `local', by which we mean the constraints are imposed 
on the nodal values of the control vector. Thus, in the constraint 
description \texttt{localconstraints.h},
these vectors are manipulated directly without additional integration. We 
note that the constraints need to be written such, that a feasible control
generates non positive entried, i.e., we 
calculate the vector
\[
 \mathcal C(q) = \begin{pmatrix}
   q_{\min} - q\\q-q_{\max}
 \end{pmatrix}.
 \]
%
Second, the \texttt{lcc} vector is used to describe the amount of unknowns that 
need to be reserved to store the constraints, and, eventually, corresponding 
Lagrange multipliers. 
Further information can be found in \texttt{basic/constraints.h}. 
In our example, we have only one block in the control (The control is stored in 
a \texttt{deal.II::BlockVector}). Hence, the \texttt{lcc} vector has a length of 
one. 
The only entry \texttt{lcc[0]} is a vector of size two (in the case of more than one 
control block, each block would be given a size of $2$). 
Each of the two entries has a specific meaning. 
The first entry \texttt{lcc[0][\textcolor{red}{0}] = 1} tells 
you how many local entries in the present block are locally constrained, here 
it is one local entry. ({\em Note that this entry will typically be one. 
it is not one  would be if we have constraints of the type $q_{i} + q_{i+1} \le 1$ for each 
even $i$, or similar combinations of multiple entries in the control vector.})

The second entry \texttt{lcc[0][\textcolor{red}{1}] = 2}, determines the number of 
constraints on this local entry. Here, we impose 
a lower and an upper bound, i.e., we give 2 constraints.
This information tells the SpaceTimeHandler, that the vector 
$\mathcal C(q)$ needs exactely twice the amount of unknowns as the vector $q$.
In general, the space needed for $\mathcal C(q)$ is given as 
\[
\frac{\text{\texttt{lcc[0][1]}}}{\text{\texttt{lcc[0][0]}}}
\]
times the unknowns for the control.

>>>>

In this example, we introduce the handling of \textit{local} constraints
whereas the mixture of local and \textit{global} constraints will be discussed
in the Example \ref{OPT_Stat_MBB-Beam}. 
%
First, we implement the upper and lower control bounds in 
\texttt{localconstraints.h}, i.e,
\[
q_{\min} \leq q \leq q_{\max}.
\]
with $q_{\min} = -500$ and $q_{\max} = 500$.
The constraints are `local', by which we mean the constraints are imposed 
on the nodal values of the control vector. Thus, in the constraint 
description \texttt{localconstraints.h},
these vectors are manipulated directly without additional integration. We 
note that the constraints need to be written such, that a feasible control
generates non positive entried, i.e., we 
calculate the vector
\[
 \mathcal C(q) = \begin{pmatrix}
   q_{\min} - q\\q-q_{\max}
 \end{pmatrix}.
 \]
%
Second, the \texttt{lcc} vector is used to describe the amount of unknowns that 
need to be reserved to store the constraints, and, eventually, corresponding 
Lagrange multipliers. 
Further information can be found in \texttt{basic/constraints.h}. 
In our example, we have only one block in the control (The control is stored in 
a \texttt{deal.II::BlockVector}). Hence, the \texttt{lcc} vector has a length of 
one. 
The only entry \texttt{lcc[0]} is a vector of size two (in the case of more than one 
control block, each block would be given a size of $2$). 
Each of the two entries has a specific meaning. 
The first entry \texttt{lcc[0][\textcolor{blue}{0}] = 1} tells 
you how many local entries in the present block are locally constrained, here 
it is one local entry. ({\em Note that this entry will typically be one. 
it is not one  would be if we have constraints of the type $q_{i} + q_{i+1} \le 1$ for each 
even $i$, or similar combinations of multiple entries in the control vector.})

The second entry \texttt{lcc[0][\textcolor{blue}{1}] = 2}, determines the number of 
constraints on this local entry. Here, we impose 
a lower and an upper bound, i.e., we give 2 constraints.
This information tells the SpaceTimeHandler, that the vector 
$\mathcal C(q)$ needs exactely twice the amount of unknowns as the vector $q$.
In general, the space needed for $\mathcal C(q)$ is given as 
\[
\frac{\text{\texttt{lcc[0][1]}}}{\text{\texttt{lcc[0][0]}}}
\]
times the unknowns for the control.
